---
title: "Survival Analysis Project"
author: "FONDZEWONG WIYFENGLA ANSLEM"
date: "2024-08-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Setup and Data Preparation

```{r}
# Clear workspace and load required libraries
rm(list = ls())
library(survival)
library(ggplot2)
library(timereg)
library(gridExtra)
library(caret)
library(survminer)
```
```{r}
# Load the TRACE dataset
data(TRACE)

# Basic checks
dim(TRACE)  # Dimensions of the dataset
head(TRACE)  # First few rows of the dataset
```



Data Summary and Initial Exploration

```{r}
# Summary statistics of the dataset
summary(TRACE)

# Count the number of unique patients
nbPatients <- length(unique(rownames(TRACE)))
paste("Number of unique patients =", nbPatients)

# Count the number of missing values
paste("Total number of NA =", sum(is.na(TRACE)))
```

Data Visualization: Histograms and Pie Charts

```{r}
# Histograms for key variables
par(mfrow = c(2, 2))
hist(TRACE$age, main = "Age Distribution", xlab = "Age")
hist(TRACE$time, main = "Time Distribution", xlab = "Time")
hist(TRACE$wmi, main = "WMI Distribution", xlab = "WMI")

# Combine status values 7 and 8 into one category for better visualization
TRACE$status2 <- ifelse(TRACE$status == 7 | TRACE$status == 8, "7&8", TRACE$status)
TRACE$status2 <- factor(TRACE$status2, levels = c(0, "7&8", 9), labels = c("Alive", "Dead other causes", "Dead from MI"))
```

```{r}
# Function to create pie charts
create_pie_chart <- function(data, labels, title) {
  data <- as.data.frame(data)
  data$labels <- labels
  data$percent <- round(data$Freq / sum(data$Freq) * 100, 1)
  data$labels <- paste(data$labels, data$percent, "%")
  
  ggplot(data, aes(x = "", y = Freq, fill = labels)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y") +
    theme_void() +
    labs(title = title) +
    geom_text(aes(label = labels), position = position_stack(vjust = 0.3), size = 2.5)
}

# Generate pie charts for categorical variables
p1 <- create_pie_chart(table(TRACE$sex), c("Male", "Female"), "Sex")
p2 <- create_pie_chart(table(TRACE$status2), c("Alive", "Dead others", "Dead from MI"), "Status")
p3 <- create_pie_chart(table(TRACE$diabetes), c("Absent", "Present"), "Diabetes")
p4 <- create_pie_chart(table(TRACE$vf), c("Absent", "Present"), "Ventricular Fibrillation")
p5 <- create_pie_chart(table(TRACE$chf), c("Absent", "Present"), "Heart Pump Failure")

# Arrange the pie charts
grid.arrange(p1, p2, p3, p4, p5, ncol = 2)

```






Nonparametric Estimation of Survival and Comparison of Groups

          This section involves Kaplan-Meier estimation, which is a nonparametric method used to estimate the survival function from the data. The overall survival probability is estimated, and the survival probability is compared between different groups, such as male and female patients. A log-rank test is performed to statistically compare the survival curves between these groups.
          


    Kaplan-Meier Estimation for Overall Survival

```{r}

km_fit <- survfit(Surv(time, status == 9) ~ 1, data = TRACE)

# Plotting
ggsurvplot(km_fit, 
           conf.int = TRUE, 
           title = "Overall Survival", 
           xlab = "Time (years)", 
           ylab = "Survival Probability")

```

After fitting the model, it generates a plot showing the survival probability over time, with confidence intervals to indicate the uncertainty around the estimated survival probabilities.



    Kaplan-Meier Estimation by Sex
    
    Here, the survival analysis is extended by grouping the data based on the sex of the individuals. This allows for a comparison of survival probabilities between different groups (in this case, males and females). A separate Kaplan-Meier curve is generated for each group, and these curves are plotted together to visually compare the survival experiences of the groups over time.
    
```{r}
km_fit_sex <- survfit(Surv(time, status == 9) ~ sex, data = TRACE)

# Plotting
ggsurvplot(km_fit_sex, 
           conf.int = TRUE, 
           title = "Survival by Sex", 
           xlab = "Time (years)", 
           ylab = "Survival Probability", 
           legend.labs = c("Male", "Female"))

```


    Log-rank test to compare survival between groups
```{r}
surv_diff <- survdiff(Surv(time, status == 9) ~ sex, data = TRACE)

# Display the result
surv_diff

```










Semi-parametric Cox Regression Analysis

    In this section we focus on building Cox proportional hazards models, which are used to assess the impact of multiple covariates on survival. Two versions of the model are created: one without interaction terms and one with interaction terms. The interaction model examines whether the effect of one variable on survival depends on another variable. The results are summarized to interpret the significance and impact of each covariate.
    

    Cox Proportional Hazards Model

```{r}
cox_model <- coxph(Surv(time, status == 9) ~ age + sex + wmi + diabetes + vf + chf, data = TRACE)
summary(cox_model)
```

    Cox Model with Interaction Terms

```{r}
cox_model_interactions <- coxph(Surv(time, status == 9) ~ wmi*chf + wmi*age + wmi*sex + wmi*diabetes + wmi*vf +
                                  chf*age + chf*sex + chf*diabetes + chf*vf +
                                  age*sex + age*diabetes + age*vf +
                                  sex*diabetes + sex*vf +
                                  diabetes*vf,
                                data = TRACE)
summary(cox_model_interactions)

```





Cox Model with Ridge Penalization

    In this section, we extend the Cox model with ridge penalization, which is a regularization technique used to handle multicollinearity and prevent overfitting. This method adds a penalty to the model coefficients, helping to shrink less important variables toward zero. The summary of this model provides insight into the effects of the covariates after applying regularization.

```{r}
fit_coxph_ridge <- coxph(Surv(time, status == 9) ~ ridge(age + sex + wmi + diabetes + vf + chf), data = TRACE)
summary(fit_coxph_ridge)
```





Cross-validation of Cox Model

    Here we implement a k-fold cross-validation process to evaluate the performance of the Cox model. The dataset is split into k subsets (folds), and the model is trained on k-1 folds while being tested on the remaining fold. This process is repeated k times, and the results are averaged to assess the model’s stability and generalizability. The cross-validation helps in understanding how well the model might perform on unseen data.
```{r}
# K-Folds Cross-validation
k = 10  # Number of folds
folds <- createFolds(TRACE$status, k = k)

# Perform cross-validation
cv_results <- lapply(folds, function(x) {
  train <- TRACE[-x,]
  test <- TRACE[x,]
  fit <- coxph(Surv(time, status == 9) ~ age + sex + wmi + diabetes + vf + chf, data = train)
  return(summary(fit))
})

# Display one of the models as an example
cv_results[[1]]
```

